Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-04-03 review.qmd
Fixed some references, removed duplicated paragraphs

A signifcant part of research into contrast processing is concerned with how these normalization mechanisms depend on colour, orientation, eye of origin, spatial and temporal frequency, age, and the presence of neurological disorders [refs: Chen and Foley, Petrov etc Carandini, Binocular normalization / suppression (Baker/Meese but also e.g. Heeger), Tadin, Winawer, Wade, Porciatti, Tsai...]. The SSVEP has proven to be invaliable in this research because it provides an objective readout of contrast representation at different parts of the visual system and allows us to 'tag' the probe and background separately.

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. For example, by varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex[@ShapleyVictor etc]. However the SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

@@

Neurons have a limited dynamic range yet they can transmit information about  visual stimuli that span many order of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear,sigmoidal contrast vs response (CRF) transducer (@Albrecht and Hamilton1982, @Tolhurst et al1981) and different neurons exhibit peak sensitivity (defined as the maximum slope of the function) at different contrast levels [@Early matteo stuff, Busse refs]. A neuronal population will therefor span a sensitivity range greater than any individual member. 
Neurons have a limited dynamic range yet they can transmit information about visual stimuli that span many orders of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear, sigmoidal contrast vs response (CRF) transducer and different neurons exhibit peak sensitivity (defined as the maximum slope of the function) at different contrast levels [@EarlyMatteoStuff; @BusseRefs]. A neuronal population will therefore span a sensitivity range greater than any individual member.

Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This "normalisation" process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed respnses of the local neuronal population.[@Heeger 92, @Carandini and Heeger 2011 @Busse 2008, @Baker and Wade]

Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This "normalisation" process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed responses of the local neuronal population [@Heeger1992; @CarandiniHeeger2011; @Busse2008; @BakerWade].

@@

Here, we show the effect of adding a second contrast component: in general, the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms --- is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan].

@@

If the two inputs are simply added together, the representation of the resulting signal in the Fourier domain is just the linear sum of the two independent signals. However, it is important to take physiology into account. If we ignore cells sensitive to contrast transients, contrast is represented as a combination of rectified signals from 'on and 'off'-sensitive cells (broadly, cells that code positive or negative changes in spatiotemporal luminance). To model the population response of these cells to a contrast-reversing luminance sine wave, we therefore introduce an initial full-wave rectification stage that simulates the outputs of two, equal populations of on- and off-polarity detectors. Once this is done, we see that the FT of the resulting signal contains power only at the second harmonics and above. It is possible to generate first-hamonic responses in contrast-response SSVEPs but only when the inputs are modulated in an 'on/off' manner rather than contrast reversed.

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code termporal changes in contrast. These cells (@ Movshon? Levitt? Earlier... Kuffler? H+W? Alpern?) will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

One of the simplest nonlinearities is the response function that describes a cell's response to different levels of contrast. In Figure xxx this is modeled as described above by a hyperbolic ratio function resulting in a saturating non-linearity. 
${R{\mathrm{max}}}\frac{C{\mathrm{in}}^n}{C{\mathrm{in}}^n+ C{\mathrm{50}}^n}$ where ${R{\mathrm{max}}}$ describes the maximum response level, $C{\mathrm{50}}$ is the 'semi-saturation constant' (the point at which the reponse is at half-maximum) and n controls the steepness of the curve.

@@

The hyperbolic ratio function is monotonic but the CRF resulting from measuring the amplitude of the 2F component contains slight roll-off  at high input contrasts. This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity. Power at other harmonics increases and the total power generated by the input is monotonic. This roll-off is often seen in experimental data and has been referred to as 'supersaturation' [@Tyler].

To illustrate the effect of contrast gain control[@Heeger92], we include a second component that contributes to the gain pool of the first. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the FT.

@@

This possibility was recognised in the early days of the technique [@reganandRegan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. Candy and Norcia (in infants) 2001, Busse et al2008, Tsai/Norcia/Wade 2010s, Then lots of Baker stuff including the CerebCort paper. : Focusing only on SSVEP: [@NorciaCandy2001, @Regan1990]

>>>>

A significant part of research into contrast processing is concerned with how these normalization mechanisms depend on colour [@Chen2000], orientation [@Foley1994], eye of origin [@Legge1979; @Baker2007], spatial and temporal frequency [@Meese2009], location [@Petrov2005; @Tadin2003; @Polat1993], age [@Betts2005], and the presence of neurological disorders [@Porciatti2000; @Tsai2011]. The SSVEP has proven to be invaliable in this research because it provides an objective readout of contrast representation at different parts of the visual system and allows us to 'tag' the probe and background at separate frequencies.

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. For example, by varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex [@Shapley1980]. However the SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

@@

Neurons have a limited dynamic range yet they can transmit information about visual stimuli that span many orders of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear, sigmoidal contrast vs response (CRF) transducer [@Albrecht1982; @Tolhurst1981] and different neurons exhibit peak sensitivity (defined as the maximum slope of the function) at different contrast levels [@Carandini1994; @Carandini1998; @Busse2009]. A neuronal population will therefor span a sensitivity range greater than any individual member.

Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This "normalisation" process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed responses of the local neuronal population [@Heeger1992; @Carandini2011; @Busse2009; @Baker2017].

@@

@@

If the two inputs are simply added together, the representation of the resulting signal in the Fourier domain is just the linear sum of the two independent signals. However, it is important to take physiology into account. If we ignore cells sensitive to contrast transients, contrast is represented as a combination of rectified signals from 'on' and 'off'-sensitive cells (broadly, cells that code positive or negative changes in spatiotemporal luminance). To model the population response of these cells to a contrast-reversing luminance sine wave, we therefore introduce an initial full-wave rectification stage that simulates the outputs of two, equal populations of on- and off-polarity detectors. Once this is done, we see that the Fourier spectrum of the resulting signal contains power only at the second harmonics and above. It is possible to generate first-hamonic responses in contrast-response SSVEPs but only when the inputs are modulated in an 'on/off' manner rather than contrast reversed.

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells (@ Movshon? Levitt? Earlier... Kuffler? H+W? Alpern?) will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

One of the simplest nonlinearities is the response function that describes a cell's response to different levels of contrast. In Figure xxx this is modeled as described above by a hyperbolic ratio function resulting in a saturating non-linearity:

\begin{equation}
R_{max} = \frac{C_{in}^n}{C_{50}^n + C_{in}^n},
\end{equation}

where ${R_{max}}$ describes the maximum response level, $C_{in}$ is the input contrast, $C_{50}$ is the 'semi-saturation constant' (the point at which the reponse is at half-maximum) and *n* controls the steepness of the curve.

@@

The hyperbolic ratio function is monotonic but the CRF resulting from measuring the amplitude of the 2F component contains a slight roll-off at high input contrasts. This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity. Power at other harmonics increases and the total power generated by the input is monotonic. This roll-off is often seen in experimental data and has been referred to as 'supersaturation' [@Tyler1985; @Peirce2007].

To illustrate the effect of contrast gain control [@Heeger1992], we include a second component that contributes to the gain pool of the first. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the FT.

@@

This possibility was recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. Candy and Norcia (in infants) 2001, Busse et al2008, Tsai/Norcia/Wade 2010s, Then lots of Baker stuff including the CerebCort paper. : Focusing only on SSVEP: [@NorciaCandy2001, @Regan1990]
